Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-02-24 paper/main.tex
improved intro.

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA automata can
be obtained for example by translating LTL formulas by
Delag~\cite{muller.17.gandalf}, ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, or
SPOT~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

These observations motivate our study of a technique reducing the
number of acceptance marks without altering the structure of the
automaton. This paper presents a technique heavily based on
\emph{quantified Boolean formulas (QBF)}. For a given TELA and
parameters $k,n$, it produces a QBF formula which is satisfiable if
and only if there exists an automaton with the same structure, $n$
acceptance marks, an acceptance formula in disjunctive normal form
with $k$ clauses, and the same set of accepting runs as the original
automaton. The placement of the marks on transitions and the
acceptance formula can be obtained from a model of the
formula. Besides this formula, we describe also the construction of
two simpler formulas such that their satisfiability implies the
existence of an automaton with the same structure, $n$ acceptance
marks, and the same set of accepting runs, but not vice versa.

@@

\paragraph*{Related results}
We are not aware about any existing research primarily aimed at
simplification of acceptance formulas of TELA or reduction of the
number of its acceptance marks without increasing the number of
states. There exists only a SAT-based approach that transforms a
deterministic TELA automaton to an equivalent automaton with a given
acceptance condition and a given number of
states~\cite{baarir.15.lpar} (if such an automaton exists). Further,
there are some SAT-based approaches aimed to reduce the number of
states of automata over infinite words. More preceisely, there are
reductions designed for nondeterministic Büchi
automata~\cite{ehlers.10.spin}, deterministic Büchi
automata~\cite{ehlers.10.sat}, and deterministic generalized Büchi
automata~\cite{baarir.14.forte}. Note that all these techniques are
usually very slow and their authors typically suggest to use them only
for specific purposes like looking for cases where some automata
construction can be improved.

@@

A transition $t=(p,a,M',q)\in\delta$ is the transition leading from
state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$. We
use $M(t)$ to refer to its set of acceptance marks $M'$. We extend
this notation to a set $T\subseteq\delta$ of transitions such that
$M(T)=\bigcup_{t\in T}M(t)$ denotes the set of acceptance marks that
appear on transitions in $T$.

\todo{Rozmyslet si, jak dal.}

@@

In graphical representation of automata, we often use acceptance marks
$\tacc{1}{1},\tacc{2}{2},\ldots\in M$ and an edge 
\begin{tikzpicture}[smallautomaton,baseline=-2pt]
  \node[state] (p) at (0,0) {$p$};
  \node[state] (q) at (1.8,0) {$q$};
  \path [->] (p) edge node[above,pos=.5,overlay] {$a$} node[accset,collacc8, pos=0.65] {k} node[accset,collacc4, pos=0.3] {j} (q);
\end{tikzpicture}
denotes the transition $(p,a,\{\tacc{4}{j},\tacc{8}{k}\},q)\in\delta$.

>>>>

TELA have attracted a lot of attention during the last five years.
Their popularity comes probably from the fact that these automata can
often use less states than equivalent automata with simpler acceptance
conditions. Further, algorithms handling TELA can automatically handle
all automata with traditional acceptance conditions. TELA automata can
be obtained for example by translating LTL formulas by
Delag~\cite{muller.17.gandalf}, ltl3tela~\cite{major.19.atva},
Rabinizer~4~\cite{kretinsky.18.cav}, or
Spot~\cite{duret.16.atva2}. There are also some algorithms processing
these automata, for example the emptiness check~\cite{baier.19.atva}
or translation of TELA to parity
automata~\cite{renkin.20.atva,casares.22.tacas}.

@@

The number of acceptance marks can be algorithmically reduced to one
as every TELA can be transformed to an equivalent Büchi automaton
(this can be easily done for example by Spot~\cite{duret.16.atva2}),
but this reduction is paid by dramatic changes of state-space: the
number of states can increase exponentially and some important
structural properties like determinism can be lost. This motivates our
study of a technique reducing the number of acceptance marks without
altering the structure of the automaton.

We present a technique heavily based on \emph{quantified Boolean
  formulas (QBF)}. For a given TELA and parameters $k,n$, it produces
a QBF formula which is satisfiable if and only if there exists an
automaton with the same structure, $n$ acceptance marks, an acceptance
formula in disjunctive normal form with $k$ clauses, and the same set
of accepting runs as the original automaton. The placement of the
marks on transitions and the acceptance formula can be obtained from a
model of the formula. Besides this formula, we describe also the
construction of two simpler formulas such that their satisfiability
implies the existence of an automaton with the same structure, $n$
acceptance marks, and the same set of accepting runs, but not vice
versa.

@@

\paragraph*{Related results}
There is a simple technique~\cite{babiak.13.spin} reducing the number
of acceptance marks in \emph{transition-based genealised Büchi
  automata (TGBA)} without changing its structure. We are not aware
about any existing research aimed at simplification of acceptance
formulas of TELA or reduction of the number of its acceptance marks
without increasing the number of states. There exists only a SAT-based
approach that transforms a deterministic TELA automaton to an
equivalent automaton with a given acceptance condition and a given
number of states~\cite{baarir.15.lpar} (if such an automaton
exists). Further, there are some SAT-based approaches aimed to reduce
the number of states of automata over infinite words. More preceisely,
there are reductions designed for nondeterministic Büchi
automata~\cite{ehlers.10.spin}, deterministic Büchi
automata~\cite{ehlers.10.sat}, and deterministic generalized Büchi
automata~\cite{baarir.14.forte}. Note that all these techniques are
usually very slow and their authors typically suggest to use them only
for specific purposes like looking for cases where some automata
construction can be improved.

@@

A tuple $t=(p,a,M',q)\in\delta$ is the \emph{transition} leading from
state $p$ to state $q$ labelled with $a$ and acceptance marks $M'$. We
use $M(t)$ to refer to the set $M'$ of acceptance marks. Further, for
a set $T\subseteq\delta$ of transitions, $M(T)=\bigcup_{t\in T}M(t)$
denotes the set of acceptance marks that appear on transitions in $T$.

@@

In graphical representation of automata, we often use acceptance marks
$\tacc{1}{1},\tacc{2}{2},\ldots\in M$. An edge 
\begin{tikzpicture}[smallautomaton,baseline=-2pt]
  \node[state] (p) at (0,0) {$p$};
  \node[state] (q) at (1.8,0) {$q$};
  \path [->] (p) edge node[above,pos=.5,overlay] {$a$} node[accset,collacc8, pos=0.65] {k} node[accset,collacc4, pos=0.3] {j} (q);
\end{tikzpicture}
denotes the transition $(p,a,\{\tacc{4}{j},\tacc{8}{k}\},q)\in\delta$.
